A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-05-05 analysis/paper/paper.Rmd
edit discussion to reply to R1's comments

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2004]. We assume that the use of imported goods and their values were increased and stressed after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts located, the remote geographical setting of northeastern Taiwan made it less interfered by European colonial power but experienced frequent trade activities stimulated by the Europeans. This paper focuses on the effect of European colonization in a pericolonial context that is not well understood compared to places that were colonial centers [@Acabado2017; @Trabert2017].

The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Network-based societies are identified in some places in Asia that were invovled in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans. Here we define trade goods as prestige goods based on their archaeological contexts where very few of trade goods were found in middens and more in burials, along with the symbol of personal influence described in the Spanish records [@LiandWu2006]. In addition, the spatial pattern of trade goods in residential presents a more concentration pattern statistically after the European presence. The evidence from different archaeological contexts indicates the possible use as prestige goods.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the major factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. We also note that no effects could be due to the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks [eg. @Coward2013, pp. 252]. We argue that the social relations at Kiwulan changed because of contact with European colonial powers and associated foreign trade. This argument is supported by changes in burial network structures through time. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person’s status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, that supports a connection between a burial and the social role of the living person. Also, ethnographic records support our interpretation of burial goods as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], so the interrelations represented by the flow of these goods observed in the burial data likely reflect social contexts of the living people at Kiwulan.

>>>>

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2004; @Cheng2017]. We assume that the use of imported goods and their values were increased and stressed after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts located, the remote geographical setting of northeastern Taiwan made it less interfered by European colonial power but experienced frequent trade activities due to local natural resources, such as rice and deerskins [@Chen2004; @Cheng2017]. This makes northeastern Taiwan a special area to explore the effect of European colonization in a pericolonial context that is still not well understood compared to European colonial bases [@Acabado2017; @Trabert2017].

The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Network-based societies are identified in some places in Asia that were invovled in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans. Here we assume that trade goods could be used as prestige goods based on their archaeological contexts where very few of trade goods were found in middens and more in burials, along with the symbol as personal influence described in the Spanish records [@LiandWu2006]. In addition, the study of spatial distribution of trade ornaments in residential at Kiwulan presents a more clustered pattern after the European presence [@LWandBM2020ornament].

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestigegoods. Our results indicate that the major factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. We also note that no effects could be due to the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

<!-- other evidence supporting the change in burial networks -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks [eg. @Coward2013, pp. 252]. We argue that the social relations at Kiwulan changed because of contact with European colonial powers and associated foreign trade. This argument is supported by changes in burial network structures through time. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person’s status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, that supports a connection between a burial and the social role of the living person. Also, ethnographic records support our interpretation of burial goods as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], so the interrelations represented by the flow of these goods observed in the burial data likely reflect social contexts of the living people at Kiwulan.

<!-- trade dynamic during this period and future work for comparison -->
We also recognize that trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have had contributed to social changes at Kiwulan. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. Historical records also support the interaction and trade activities between Europeans and local Indigenous people. The trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at Kiwulan were frequently found from archaeological sites with Dutch occupation, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides an ideal example to explore the reaction of Indigenous people to European colonial presence at a settlement scale. We still need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence from the periphery with colonial bases, such as Helping Dao and Tamsui. However, Indigenous sites related to the 17th century with intensive excavation and sufficient evidence are rare to offer good comparison currently and future work is still needed.
